---
title: "ps7 solutions"
output:
  pdf_document: default
  html_document: default
---

# Question 2

## prelims
```{r}
rm(list = ls())
#LOAD DATA
data = read.csv("~/Desktop/MetabolicRate.csv")
dataBody = data$BodySize
dataInstar = data$Instar
dataMrate = data$Mrate

nObs = length(dataBody)
print(nObs)

#LOG DATA
dataBody = log(dataBody)
dataMrate = log(dataMrate)

setwd("~/Desktop")

library(rstan)
#library(ggplot2)
options(mc.cores = parallel::detectCores())
```

## Step 2
```{r}
#RUN THE MODEL WITH 4 CHAINS
fit = stan('PS_7.stan', iter = 25000, chains = 4,
           data = list(y=dataMrate, x1=dataBody, x2=dataInstar, N = nObs))

print(fit,probs = c(0.25,0.5,0.75))
```
the esitmated values match those from problem set 4. 

## step 3
```{r}
#TRACEPLOT
traceplot(fit, window=c(23000, 25000))
```
The traceplots show good mixing among the four chains and they are oscillating 
about the stationary distribution. Therefore, there aren't any errors in the 
stan file. 

## step 4
```{r}
#MATRIX SCATTERPLOTS
pairs(fit, pars = c("beta0", "beta1", "beta2", "sigma"))
```

## step 6
```{r}
fit = stan('ps_7_2.stan', iter = 10000, chains = 4,
           data = list(y=dataMrate, x1=dataBody, x2=dataInstar, N = nObs))

#print(fit,probs = c(0.25,0.5,0.75))

library(bayesplot)
ypreds = extract(fit)[["yPred"]]
len = dim(ypreds)[1]
print(len)
matrix = ypreds[floor(runif(100) * len), ]
#print(matrix)

ppc_dens_overlay(
  dataMrate,
  matrix,
  size = 0.25,
  alpha = 0.7,
  trim = FALSE,
  bw = "nrd0",
  adjust = 1,
  kernel = "gaussian",
  n_dens = 1024
)
```
By looking at the plot, we see that the predicted distributions (yrep) don't line 
up with the data (y) very well on the interval [-0.5, 3.5]. Therefore, we conclude that the linear model is a decent model (because the curves have similar shapes), but doesn't provide a very accurate or satisfactory prediction (because the curves don't overlap well on the interval). The true model is probably more complicated than this linear one.

## step 7
```{r}
library(ggplot2)
print(max(dataInstar))

# need to separate out the y=dataMrate and the columns of the matrix
I1 = c()
I2 = c()
I3 = c()
I4 = c()
I5 = c()

#append value n \in [1, 305] to corresponding Ij vector
for (n in 1:nObs) {
  if (dataInstar[n] == 1) {
    I1 = append(I1, n)
  } else if (dataInstar[n] == 2) {
    I2 = append(I2, n)
  } else if (dataInstar[n] == 3) {
    I3 = append(I3, n)
  } else if (dataInstar[n] == 4) {
    I4 = append(I4, n)
  } else if (dataInstar[n] == 5) {
    I5 = append(I5, n)
  }
}

#make sub-matrices for each j=1:5
matrix1 = matrix[ , I1]
matrix2 = matrix[ , I2]
matrix3 = matrix[ , I3]
matrix4 = matrix[ , I4]
matrix5 = matrix[ , I5]

#make sub y-vectors for each j=1:5
y1 = dataMrate[I1]
y2 = dataMrate[I2]
y3 = dataMrate[I3]
y4 = dataMrate[I4]
y5 = dataMrate[I5]

#make one overlay plot for each j=1:5
p1 = ppc_dens_overlay(
  y1,
  matrix1,
  size = 0.25,
  alpha = 0.7,
  trim = FALSE,
  bw = "nrd0",
  adjust = 1,
  kernel = "gaussian",
  n_dens = 1024
)
p1 + ggtitle("Instar = 1")

p2 = ppc_dens_overlay(
  y2,
  matrix2,
  size = 0.25,
  alpha = 0.7,
  trim = FALSE,
  bw = "nrd0",
  adjust = 1,
  kernel = "gaussian",
  n_dens = 1024
)
p2 + ggtitle("Instar = 2")

p3 = ppc_dens_overlay(
  y3,
  matrix3,
  size = 0.25,
  alpha = 0.7,
  trim = FALSE,
  bw = "nrd0",
  adjust = 1,
  kernel = "gaussian",
  n_dens = 1024
)
p3 + ggtitle("Instar = 3")

p4 = ppc_dens_overlay(
  y4,
  matrix4,
  size = 0.25,
  alpha = 0.7,
  trim = FALSE,
  bw = "nrd0",
  adjust = 1,
  kernel = "gaussian",
  n_dens = 1024
)
p4 + ggtitle("Instar = 4")

p5 = ppc_dens_overlay(
  y5,
  matrix5,
  size = 0.25,
  alpha = 0.7,
  trim = FALSE,
  bw = "nrd0",
  adjust = 1,
  kernel = "gaussian",
  n_dens = 1024
)

p5 + ggtitle("Instar = 5")
```

The model seems to predict data best for Instar = 3. The predictions are decent for Instar=1, 2, and 4, and fairly bad for Instar=5.  This may be because there isn't much data for Instar=5 compared with the other values. The other inconsistencies may be due to the limitatins of the linear model. 
Also, perhaps the relationship between body size and metabolic rate changes with Instar value and thus we would need a different model for each one. 
